A hierarchical approach for modelling X-ray beamlines. Application to a coherent beamline.

### 2019-03-29 paper-hierarchical.tex
Updates from Overleaf

After the analysis of how the ID16A optics modify the characteristics of the x-ray beam done in the previous section, when one looks back to the numbers obtained by simple analytical canculations, one can verify that these analytical previsions were a quite good approximation of the values then obtained using more sophisticated methods. The size of the focused beam and the transmission calculated using ray tracing are in good agreement.

When comparing beam sizes, one usually assume that the intensity distributions are Gaussian and therefore well characterised by the FWHM value. From Fig.~\ref{fig:ray tracing} one can see that the horizontal intensity distribution is not Gaussian for the ESRF, as a result of cropping the beam with the VSS opened to 50$\mu$m. Although the electron beam statistics is in a very good approximations Gaussian, the undulator emission is not. In our ray tracing simulations we always suppose that we work at the resonance, then we applied the Gaussian approximation used in Eqs.~\ref{eq:photon small sigmas} and \ref{eq:photon big sigmas} to compute the geometrical characteristics of the source. This is certainly acceptable at the resonance but will fail out of resonance. Indeed, these lack of Gaussianity is relevant in the calculations of the coherent fraction. We approximate the coherence fraction by Eq.~\ref{eq:coherent fraction}, which for zero emittance ($\sigma_{x,y}=\sigma_{x',y'}$=0) becomes CF=1, as expected because a zero emittance electron beam, that is a single electron or a filament beam affected by an magnetic field emmits a fully coherent wavefront. However, the photon source for a filament beam would have an emittance of $\sigma_u \sigma_{u'} \approx \lambda / 2 \pi$ (using Eqs.~\ref{eq:photon small sigmas}), which is in contradiction with theory, that says that a pure coherent state has emittance of $\lambda / 4 \pi$, as happens for a Gaussian photon beam. This is justified considering that Eqs.~\ref{eq:photon small sigmas} are approximations obtained by fitting with Gaussians the theoretical non-Gaussian distributions (see Ref.~\cite{elleaume}). The correct definition of CF as the occupation of the lowest coherent mode does not present these incongruences. However, its calculation requires to perform the coherent mode decomposition, as done in Section~\ref{comsyl} which is very costly. It is therefore the case to find the compromise between accuracy and cost, or where in the hierarchical series of methodologies proposed one should stop.

The combination of ray tracing and wave optics methods used indicate that the focalized beam size due to  demagnification with mirrors (giving for EBS 10x5 nm2 analytically, 8.4x4.8 nm2 with ray tracing) has to be corrected with diffraction effect (8.4x7.8 hybrid, 12.7x9.2 SRW) and then add the slope error effect. The slope errors are very small, necessary not to blow up the size and kit it in the specifications range (10.8x8.2 hybrid 13.6x9.9 SRW). It is important to treat the the slope errors correctly. Slope errors can be included in the ray tracing using the data in Appendix A and applying specular reflection on the optical surface with errors. This is the standard mode of including slope errors in SHADOW, using measured profiles or synthetic ones. It works well for the case of incoherent beams, but in our case, because of the high coherence of the beam the ray tracing method (specular reflection) overestimates the blowing of the focal spot conducting to wrong results. Therefore the slope errors of the KB system have been considered using a ``wave optics approach'', as explained in Ref.~\cite{hybrid}. Here, the mirror heigh errors $z(x)$ are projected to the plane perpendicular to the propagation ($z(x) \sin \theta$, with $\theta$ the grazing angle) and originate a phase shift $\psi$ in the wavefront phase proportional to the optical path $\psi = - k z(x) \sin \theta $. This is the algorithm used in the simulations in Sections~\ref{level1}, \ref{srw_se} and \ref{srw_me}.

@@

The wavefront simulation of the beamline include propagation of the beam from element to element (drift spaces) using different possible propagators. A propagator based on the Freskel-Kirchhoff integral is used. The applications of this (and other) operators imply the calculation of an integral (or sum) at each pixel of the image plane. This is costly from the computer point of view, so Fourier Transforms (Fourier Optics) are used to reduce complexity and calculation time. This is practically the only choice for simulating 2D images. The use of Fourier Transforms adds another problem: the result is very dependent on the sampling of the wavefront (i.e., pixel size, number of points, dimension of the window where the wavefront is sample). This means the usually one has to test many configurations before finding one that works reaonably well (even though one can never guarantee the complete accuracy of the sampling used). A full wavefront optics simulation is an iterative process, where the user has to refine the sampling parameters element by element to be sure the wavefront at each position is sampled correctly. This is a time consuming work for performing it with the full 2D simulation. For this reason, we found very useful to start a wavefront calculation in 1D, separating the horizontal and vertical simulations, and using a simplified system with ideal sources and ideal focusing elements. This is ofted a good choice because the horizontal and vertical coupling is usually small in synchrotron beamlines. This 1D simulations help in refining the choice of propagator and its parameters (resampling, zoom factors) that will be valid if a, more costly, full 2D simulation is envisaged.

Many of the simulations presented here (ray tracing and coherent optics in Sections XXXXX) can be done interactively in a normal laptop. However, when one one to obtain information on the partial coherence, very long calculations have to be done. For example, Monte Carlo calculations in Section~\ref{srw_me} took significantly longer to be completed for the multi electron case: a few hours on a 56-core computer cluster. For performing the coherent mode decomposition the calculation time depends very much on the dimensioning on the problem. The EBS case is reasonable: we sampled the wavefronts using 1007x335 pixels and calculated 1103 modes that represent 98\% of the spectral density. This represents 5 MB per wavefront and a matrix of 1695 GB to diagonalize. It took less that four hours in the a cluster of 28 cores. The calculation is more expensive if one wants to decompose a source that is pretty incoherent like our ESRF-High-$\beta$. This is a limit case, as the wavefields have to be vey large XXxXX to accomodate the large espatial extension of the source. The calculation presented here took XX days in a large XX cluster. To give an idea of the real cost, a similar work send to the AWS cluster costed about 2000 USD. This was done as a test case, to evaluate the limits of the calculations, but it is certainly not recommended to do similar things, not only for the cost but also for the lack of interest to treat with partial coherence methods a source that is quite incoherent. In Appendix B we present for the ESRF High$\beta$ similar results than those in Section~\ref{comsyl}.

There are some effects that are typically studied when designing and simulating a beamline that have been ignores here. For example the study of the power emitted by the source and its heat load effect in the first (or firsts) slits and optical elements (here the ML monochromator). Also reflectivity of the elements are not considered. Thir effect can also be fully simulated, but we preferred here to integrate them in the simulations for the sake of clarity. Again an analytical estimation would give a reflectivity about 0.7 for multilayers and 0.9 for mirrors, therefore a reduction factor of (0.7)**3 ML for the beamline flux due to reflectivity of mirror coatings.

In conclusion, we have shown how to obtain the relevant parameters necessary to quantify flux, transmission, fucused beam size, and coherent fraction for a modern beamline with already exceptional performances in the old ESRF DBA lattices, and advanced the improvements to be expected when the EBS will start in 2020. In this discussion we obtained these parameters using a hierarchical flowchart, with increasing complexity. The phase of analytical estimation of the main parameters is essential, not only to start with some values but also because these values help to perform sanity check of the computer simulations. A ray tracing simulation is inexpensive and always recommended. The comparison of the CF beamline transmittivity would give an idea of how coherent is the beam, or how many modes will be present in your final beam. For high coherence (a few modes), like for ID16A, the wave optics calculations are necessary. If the final beam has a very high coherence, like in our case, coherent optics (1D simplifications, and single electron calculations) may give good results. We may have stopped there, as in our case coherence is very high. The COMSYL calculations presented here introduce a new way of quantitatively assess the quality of the partial coherent beam via the calculation of the coherent fraction.

@@

 %-------------------------------------------------------------------------
 % Manuel & Rafael
 %-------------------------------------------------------------------------
\section{Summary}

Bla bla

>>>>

After the complete analysis done in the previous section on how the ID16A optics modify the characteristics of the x-ray beam, one can verify that the analytical previsions were a good estimator of the values then obtained using more sophisticated methods. The size of the focused beam (e.g, 10x5 for EBS) and the transmission (3\%) calculated analytically are in good agreement with geometric ray tracing (8.4x4.8, and 2.28\%). When characterising beam sizes, one usually assume that the intensity distributions are Gaussian and therefore they are well characterised by the FWHM value. From Fig.~\ref{fig:ray tracing} one can see that the horizontal intensity distribution is not Gaussian for the ESRF, as a result of cropping the beam with the VSS opened to 50$\mu$m. Although the electron beam statistics is in a very good approximations Gaussian, the undulator emission is not. In our ray tracing simulations we always suppose that we work at the resonance, then we applied the Gaussian approximation used in Eqs.~\ref{eq:photon small sigmas} and \ref{eq:photon big sigmas} to compute the geometrical characteristics of the source. This is certainly acceptable at the resonance but will fail out of resonance. Indeed, these lack of Gaussianity is relevant in the calculations of the coherent fraction. We approximate the coherence fraction by Eq.~\ref{eq:coherent fraction}, which for zero emittance ($\sigma_{x,y}=\sigma_{x',y'}$=0) becomes CF=1, as expected because a zero emittance electron beam (a single electron or a filament beam) that travel in an magnetic field emits a fully coherent wavefront. However, the photon source for a filament beam would have an emittance of $\sigma_u \sigma_{u'} \approx \lambda / 2 \pi$ (using Eqs.~\ref{eq:photon small sigmas}), which is in contradiction with theory, that says that a pure coherent state has emittance of $\lambda / 4 \pi$, as happens for a Gaussian photon beam. This is justified considering that Eqs.~\ref{eq:photon small sigmas} are approximations obtained by fitting with Gaussians the theoretical non-Gaussian distributions (see Ref.~\cite{elleaume}). The correct definition of CF as the occupation of the lowest coherent mode does not present these incongruences. However, its calculation requires to perform the coherent mode decomposition, as done in Section~\ref{comsyl} which is very costly. It is therefore the case to find the compromise between accuracy and cost, or where in the hierarchical series of methodologies proposed one should stop.

The combination of ray tracing and wave optics methods used indicate that the focalized beam size due to  demagnification with mirrors (giving for EBS 10x5 nm2 analytically, 8.4x4.8 nm2 with ray tracing) has to be corrected with diffraction effect (8.4x7.8 hybrid, 12.7x9.2 SRW) and then add the slope error effect. The slope errors are very small, necessary not to blow up the size and kit it in the specifications range (10.8x8.2 hybrid 13.6x9.9 SRW). It is important to treat the the slope errors correctly. Slope errors can be included in the ray tracing using the data in Appendix A and applying specular reflection on the optical surface with errors. This is the standard mode of including slope errors in SHADOW, using measured profiles or synthetic ones. It works well for the case of incoherent beams, but in our case, because of the high coherence of the beam the ray tracing method (specular reflection) overestimates the blowing of the focal spot conducting to wrong results. Therefore the slope errors of the KB system have been considered using a ``wave optics approach'', as explained in Ref.~\cite{hybrid}. Here, the mirror height errors $z(x)$ are projected to the plane perpendicular to the propagation ($z(x) \sin \theta$, with $\theta$ the grazing angle). They introduce a phase shift $\psi$ in the wavefront phase proportional to the optical path $\psi = - k z(x) \sin \theta $. This is the algorithm used in the simulations in Sections~\ref{level1}, \ref{srw_se} and \ref{srw_me}.

@@

The wavefront simulation of the beamline include propagation of the beam from element to element (drift spaces) using different possible propagators. A propagator based on the Freskel-Kirchhoff integral is used. The applications of this (and other) operators imply the calculation of an integral (or sum) at each pixel of the image plane. This is costly from the computer point of view, so Fourier Transforms (Fourier Optics) and its Fast Fourier Transform implementation are used to reduce amount of operations and the calculation time. This is practically the only choice for propagating 2D wavefronts. The use of Fourier Transforms adds another problem: the result is very dependent on the sampling of the wavefront (i.e., pixel size, number of points, dimension of the window where the wavefront is defined). This means the usually one has to test many configurations before finding one that works reasonably well (even though one can never guarantee the complete accuracy of the sampling used). A full wavefront optics simulation is an iterative process, where the user has to refine the sampling parameters element by element to be sure the wavefront at each position is sampled correctly. This is a time consuming work in particular for 2D simulation. We found very useful to start a wavefront simulation in 1D, separating the horizontal and vertical simulations, using a simplified system with ideal sources and ideal focusing elements. The use of 1D propagation is justified because in in synchrotron beamlines the horizontal and vertical coupling is usually small. This 1D simulations help in refining the choice of propagator and its parameters (resampling, zoom factors) that can be reused in a full 2D simulations.

Regarding computational effort, simulations concerning ray tracing (Section~\ref{level1}) and coherent optics (Sections~\ref{wofry} and \ref{srw_se}) can be done interactively in a normal laptop. However, for those concerning partial coherence, very long calculations are needed. For example, Monte Carlo calculations in Section~\ref{srw_me} for the multi electron case \inred{HOW MANY ELECTRONS?} spent a few hours on a 56-core computer cluster. For performing the COMSYL coherent mode decomposition the calculation time depends very much on the dimensions on the problem. For EBS the wavefronts were sampled using 1007x335 pixels, and 1103 modes were computed representing 98\% of the spectral density. This represents 5 MB per wavefront and a matrix of 1695 GB to diagonalize. It took less that four hours in the a cluster of 28 cores. The calculation is more expensive if one wants to decompose a source that is pretty incoherent like our ESRF-High-$\beta$. This is a limit case, it was used wavefields of XxX pixels to accommodate the large spatial extension of the source. The calculation took XX days in a large XX cluster. 
%To give an idea of the real cost, a similar work send to the AWS cluster costed about 2000 USD. 
This was done as a test case, to demonstrate the feasibility of the calculations. It is not recommended to do similar calculations that imply a the cost because a source that is quite incoherent was treated with partial coherence methods. In Appendix B we present for the ESRF High$\beta$ similar results than those in Section~\ref{comsyl}.

Some effects that are typically studied when designing and simulating a beamline have not been discussed here. For example the study of the power emitted by the source and its heat load effect in the first (or firsts) slits and optical elements (here the ML monochromator). Also, the reflectivity of the elements are not considered. The simulation of this was not included in the simulations, because it can be estimated analytically considering an average reflectivity about \inred{0.7 for multilayers and 0.9 for mirrors, therefore a reduction factor of (0.7)**3 ML for the beamline flux due to reflectivity of mirror coatings}.

\todo{discuss coherence length}

\section{Summary and conclusions}

We have obtained quantitative values of flux, transmission, size of the focused beam, and coherent fraction for ID16A, a modern beamline with already exceptional performances in terms of nanofucusing and coherence. We predicted advances the improvements in these parameters when the EBS will start in 2020. We obtained these parameters using a hierarchical flowchart with, applying methods with increasing complexity and requiring higher computational effort.

The analytical estimation of the main parameters is mandatory first step in the beamline conception and analysis. We applied concepts based on optical magnification and numerical aperture acceptance and calculated approximately the coherent fraction at the source. These values help to perform sanity check of the computer simulations. A ray tracing simulation is inexpensive and provides important information. The comparison of the CF with the beamline transmitivity would give an idea of the beam coherence, reflected in the number modes present in the final beam. For high coherence, like for ID16A, the wave optics calculations are necessary. If the final beam has a very high coherence, like in our case, coherent optics (1D simplifications, and single electron calculations) may give good results. Full calculations based on coherent mode decomposition Were introduced here. It is a new way of quantitatively assess the quality of the partial coherent beam via the calculation of the coherent modes of the source, propagation of the modes and recalculating the coherent mode expansion on the final beam from which one obtains the CF, a parameter that valutated the coherence quality.

@@
